Clinical advancement forecasting \thanks{\textit{\underline{Citation}}: \textbf{TBD}}

### 2024-04-08 paper/paper.tex
Refactor introduction

It has been well established that drugs with human genetic evidence linking their respective targets to indications in clinical trials are more likely to succeed \cite{Nelson2015-eg,King2019-rc,Minikel2023.06.23.23291765,Razuvayevskaya2023.02.07.23285407,PMID:30652614,PMID:24833294,PMID:35804044,PMID:37803084,PMID:36963162}, and to a lesser extent, that the same may also be true for single-cell transcriptomic evidence \cite{Dann2024.04.04.24305313}. This information has been used to devise many target and target-disease ranking algorithms based primarily on a synthesis of multiple genetic signals alone \cite{PMID:38172303,Koscielny2017-rr,PMID:31253980}. It is also possible to expand the breadth of this genetic support to more targets/diseases based on knowledge graphs, protein interactions and/or disease ontologies \cite{PMID:33262371,Bao2022-bq,Sadler2023-xd,PMID:36087372,PMID:36823319}. To our knowledge, all such expansion methods identify a larger space of opportunities at the expense of expected success rates. This is not a focus of this work as we aim, instead, to establish a framework for identifying targets/diseases with the very highest possible likelihood of success first. This is accomplished by integrating human clinical, genetic/genomic, transcriptomic and proteomic data as well as cell/animal model evidence, pathway information and basic literature metrics from Open Targets \cite{Koscielny2017-rr} in a simple statistical modeling framework. It can be contrasted with far more integrative methods that rely on neural and/or graph models over extensive knowledge graphs \cite{Paliwal2020-hr,PMID:33741907,pittala2020relationweighted,PMID:32750869}, which are more complex and difficult to interpret. We believe a desirable middle ground between these approaches and those that aim to combine many orthogonal indicators of success through expert knowledge in heuristic systems \cite{PMID:38404138,Koscielny2017-rr} would: 1) permit inclusion of many types of evidence from many sources, 2) be highly interpretable, 3) support expert judgement where necessary, and 4) not require manual ranking/weighting schemes.

A substantial challenge inherent to building such a system is the need to account for the longitudinal nature of knowledge discovery in biomedicine. This is vital because any method that optimizes for likely future clinical success based on historical clinical success may easily be biased by the non-random nature with which evidence is absent otherwise. We use only "temporalized" evidence, i.e. evidence for which timing of its emergence can be determined, and outcomes when training and evaluating our methods before ultimately applying them to present-day evidence with no restrictions on timing. We discuss motivations, prior research and our own analysis on how important this problem is for each source in Section~\ref{sec:results_inflation}.

Addressing such problems is common, but not ubiquitous, in studies that attempt to predict clinical trial outcomes using temporalized predictors \cite{PMID:37483175, PMID:34430930, Lo2019Machine}. The need for this is often clear in that setting where the inclusion of predictors like historical success rates for targets/diseases, trial sponsor track records, eventual patient enrollment, etc. constitute clear information leaks otherwise. This is discussed more in \cite{PMID:37483175} which notes several studies that do not account for this problem, and presents "quasi-prospective" as well as true "prospective" results. The difference between the two is that the former reconstructs timelines for predictors and outcomes based on recorded event dates while the latter relies on frozen predictions that are never evaluated until years later. Nomenclature for these formulations is conflicting though, where this definition of a "quasi-prospective" design is deemed entirely prospective in some cases, e.g. \cite{PMID:37225853}. We will refer to our design in this study as quasi-prospective since this definition is the best fit.

The prior works discussed so far can largely be categorized as either 1) target and target-disease prioritization methods evaluated based on how well they correlate with clinical trial success and 2) clinical trial outcome prediction models. Both are measured against the same outcomes and an important distinction between them lies in how the former methods are {\bf not} directly optimized for those outcomes while the latter methods are. In this study, we attempt to bridge these methodologies by predicting clinical trial advancement for target-disease pairs based solely on information that would be present well in advance of any drug program or individual trial. We then calibrate these predictions to determine what thresholds are necessary to match the observed success rates from benchmarks for genetic support like OMIM \cite{PMID:15608251}, ClinVar \cite{PMID:24234437} and GWAS. Finally, we examine how many present-day target-disease pairs are undeveloped (i.e. have never been in clinical trials), have a tractable target and are likely to see success rates matching or exceeding those calibrated benchmarks.

@@

In order to model clinical advancement for target-disease (TD) pairs, we first define "advancement" as progression beyond any particular trial phase across all drugs associated with any one TD pair as indicated by the presence of a later-stage trial. All results to follow consider only advancement beyond phase 2 due to limitations described in Section~\ref{sec:discussion}. This binary outcome is then predicted based on a list of features shown in Table~\ref{tab:features}. Information for each of these features is only used when it was published before the year \textbf{prior} to the first phase 2 trial observed, with an exception for genetic evidence discussed in Section~\ref{sec:results_inflation}. A training dataset is then formed by including only TD pairs where this first phase 2 year is between 1990 and 2015. The evaluation dataset then consists of all TD pairs entering phase 2 between 2016 and 2022, with a 2 year offset from the present year (2024) to allow enough time for some trials to complete. While the average phase 2 trial duration may be as low as 2 years \cite{fdaStepClinical}, other estimates would suggest half of them take longer than 2.9 years \cite{PMID:29394327}. This means a substantial fraction of outcomes are censored, that this is an important parameter to test sensitivity to and that time itself is likely to be a crucial covariate in this formulation. The distribution of these outcomes, the number of associated targets/diseases and a variety of other statistics on this dataset are presented in Supplementary~Figure~\ref{fig:dataset_statistics}.

>>>>

It has been well established that drugs with human genetic evidence linking their respective targets to indications in clinical trials are more likely to succeed \cite{Nelson2015-eg,King2019-rc,Minikel2023.06.23.23291765,Razuvayevskaya2023.02.07.23285407,PMID:30652614,PMID:24833294,PMID:35804044,PMID:37803084,PMID:36963162}, and to a lesser extent, that the same may also be true for single-cell transcriptomic evidence \cite{Dann2024.04.04.24305313}. This information has been used to devise target and target-disease ranking algorithms based primarily on a synthesis of multiple genetic signals alone \cite{PMID:38172303,Koscielny2017-rr,PMID:31253980}. It is also possible to expand the breadth of this genetic support to more targets and diseases based on knowledge graphs, protein interactions and/or disease ontologies \cite{PMID:33262371,Bao2022-bq,Sadler2023-xd,PMID:36087372,PMID:36823319}. To our knowledge, all such expansion methods identify a larger space of opportunities at the expense of expected success rates. This is not a focus of this work as we aim, instead, to establish a framework for identifying target-disease (TD) pairs with the very highest possible likelihood of success. We accomplish this by integrating human clinical, genetic/genomic, transcriptomic and proteomic data as well as cell/animal model evidence, pathway information and basic literature metrics from Open Targets \cite{Koscielny2017-rr} in statistical models that are trained on past clinical outcomes, evaluated primarily on their performance at upper extremes of predictions and then deployed on undeveloped TD pairs to establish an evidence-based ranking pipeline. This method can be contrasted with far more integrative methods that rely on neural and/or graph models over extensive knowledge graphs \cite{Paliwal2020-hr,PMID:33741907,pittala2020relationweighted,PMID:32750869}, which are more complex and difficult to interpret. We believe a desirable middle ground between these approaches and those that aim to combine many orthogonal indicators of success through expert knowledge in heuristic systems \cite{PMID:38404138,Koscielny2017-rr} would: 1) permit inclusion of many types of evidence, spanning hundreds or even thousands of putative orthogonal, causal predictors, 2) be highly interpretable, 3) support expert judgement where necessary, and 4) not require manual ranking/weighting schemes. This is the basis for our internal ranking platform, and such operational advantages have made it possible to reach or exceed the inclusivity of highly integrative methods without sacrificing interpretability, modularity and ease of deployment.

A substantial challenge inherent to building such a system is the need to account for the longitudinal nature of knowledge discovery in biomedicine. This is vital because any method that optimizes for likely \textit{future} clinical success based on \textit{historical} clinical outcomes may easily be biased by the non-random nature with which evidence can be absent otherwise. We use longitudinal evidence, i.e. evidence for which timing of its emergence can be determined, where appropriate when training and evaluating our methods before ultimately applying them to present-day evidence that is not temporalized. We discuss motivations, prior research and our own analysis on how important this problem is for each type of evidence in Section~\ref{sec:results_inflation}.

While the use of longitudinal evidence is very rare in studies on target prioritization algorithms, it is not at nearly as rare in studies that attempt to predict clinical trial outcomes using a variety of biological evidence sources, clinical statistics and other organizational or behavioral factors \cite{PMID:37483175, PMID:34430930, Lo2019Machine}. The need for this is often clear in that setting where the inclusion of predictors like historical success rates for targets or diseases, trial sponsor track records, eventual patient enrollment, etc. constitute clear information leaks otherwise. This is discussed in some depth in \cite{PMID:37483175} which notes several studies that do not account for this problem before drawing a clear distinction between quasi-prospective and prospective problem formulations. The difference between the two is that the former reconstructs timelines for predictors and outcomes based on recorded event dates while the latter relies on frozen predictions that are never evaluated until sufficient time has elapsed for more outcomes to occur. Nomenclature for these formulations is conflicting though, where this definition of a quasi-prospective formulation is deemed entirely prospective in some cases, e.g. \cite{PMID:37225853}. To avoid any potential confusion, our formulation in this study is quasi-prospective. The distinct advantages and disadvantages of each formulation provide a complementary risk/reward trade-off that we capitalize on in our internal platform, and we assert that the distinction between them is important. Interestingly, the existence of Open Targets snapshots dating back to at least 2016 would enable the extension of this work to a prospective formulation as well. This is discussed more in Section~\ref{sec:discussion}.

The preceding works discussed so far can largely be categorized as either 1) target and target-disease prioritization methods evaluated based on how well they correlate with observed clinical trial success and 2) clinical trial outcome prediction models. Both are measured against the same outcomes and an important distinction between them lies in how the prioritization methods are {\bf not} directly optimized for those outcomes while the outcome prediction methods are. In this study, we attempt to bridge these methodologies by predicting clinical trial advancement for target-disease pairs based solely on information that could be present well in advance of any drug program or individual trial. We then calibrate these predictions to determine what thresholds are necessary to match the observed success rates from benchmarks for genetic support like OMIM \cite{PMID:15608251}, ClinVar \cite{PMID:24234437} and GWAS. Finally, we examine how many present-day target-disease pairs are undeveloped (i.e. have never been in clinical trials), have a tractable target and are likely to see success rates matching or exceeding those benchmarks.

@@

In order to model clinical advancement for target-disease pairs, we first define "advancement" as progression beyond any particular trial phase across all drugs associated with any one TD pair as indicated by the presence of a later-stage trial. All results to follow consider only advancement beyond phase 2 due to limitations described in Section~\ref{sec:discussion}. This binary outcome is then predicted based on a list of features shown in Table~\ref{tab:features}. Information for each of these features is only used when it was published before the year \textbf{prior} to the first phase 2 trial observed, with an exception for genetic evidence discussed in Section~\ref{sec:results_inflation}. A training dataset is then formed by including only TD pairs where this first phase 2 year is between 1990 and 2015. The evaluation dataset then consists of all TD pairs entering phase 2 between 2016 and 2022, with a 2 year offset from the present year (2024) to allow enough time for some trials to complete. While the average phase 2 trial duration may be as low as 2 years \cite{fdaStepClinical}, other estimates would suggest half of them take longer than 2.9 years \cite{PMID:29394327}. This means a substantial fraction of outcomes are censored, that this is an important parameter to test sensitivity to and that time itself is likely to be a crucial covariate in this formulation. The distribution of these outcomes, the number of associated targets/diseases and a variety of other statistics on this dataset are presented in Supplementary~Figure~\ref{fig:dataset_statistics}.
